---
title: "M3 Lab"
author: "Matthew Præstgaard"
date: "due 10/23/2022"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "dark"
    downcute_theme: "default"
    code_folding: "show"
  svglite:
    fig.retina: 2
---

```{r,echo=FALSE,message=FALSE,warning=FALSE}

require(knitr)
knitr::opts_chunk$set(error = TRUE)

```

# Part 0: Loading and Cleanup

## Housekeeping

### Libraries

```{r, message = FALSE}
library(tidyverse)
library(readxl)
library(httpgd)
library(ggthemr)
ggthemr("flat dark")
library(markdown)
library(rmdformats)
library(plotly)
library(moments)
library(gridExtra)
library(ggh4x)

```

## Functions
### I just think it's kind of fun to write them 

```{r}

#testing a histogram with summary stats included, my magnum opus function
#x column must be passed as a string i.e. "column" until I become better at writing functions
#usage: histogramCurveSum(data, "x", title = "title" (optional), x_title = "x_title" (optional))
hist.summarycurve <- function(data, x, title = "Insert Title", x_title = "Insert x-Title") {
  require(tidyverse)
  require(ggthemr)
  require(moments)
  require(gridExtra)

  ggthemr("flat dark")

  # Calculate the summary statistics
  mean_val <- mean(data[[x]])
  sd_val <- sd(data[[x]])
  median_val <- median(data[[x]])
  iqr_val <- IQR(data[[x]])
  skew_val <- moments::skewness(data[[x]])

  # Create a summary table
  summary_table <- data.frame(
    Statistic = c("Mean", "SD", "Median", "IQR", "Skewness"),
    Value = c(mean_val, sd_val, median_val, iqr_val, skew_val)
  )

  # Create the histogram plot
  histogram_plot <- ggplot(data, aes_string(x = x)) +
    geom_histogram(
      aes(y = after_stat(density)),
      bins = 50,
      color = "#0052eb") +
    stat_function(
      fun = dnorm,
      args = list(mean = mean_val, sd = sd_val),
      size = 1,
      linetype = "dashed",
      col = "#ecd467"
    ) +
    labs(
      title = title,
      x = x_title
    )

  # Combine the histogram plot with the summary table
  grid.arrange(histogram_plot, tableGrob(summary_table), ncol = 2, widths = c(3, 1))
}

#faceted histograms by grouping variable with density curves
#usage: hist.dodge(data, data$x, data$group, "title")
#must specify data$arg since [[x]] and {{x}} don't work for me most of the time
hist.dodge <- function(data, x, group, title = "Insert Title") {

  require(ggplot2)
  require(ggh4x)

  ggplot(data, aes(x = x, fill = group)) +
    geom_histogram(aes(
      y = after_stat(density)),
      bins = 45,
      alpha = 0.8,
      color = "#353535") +
    stat_theodensity(
      col = "#ecd467",
      size = 0.9,
      linetype = "dashed"
    ) +
    labs(
      title = title,
      x = "Group",
      y = "Frequency"
    ) +
    facet_wrap(group)

}

#function to summarize variables without retyping stuff
#the function automatically drops NA values
#usage: summarize_variable_drop(dataset, var)
summarize.var <- function(dataset, var) {

require(tidyverse)

  dataset %>%
    dplyr::summarise(
      mean = mean({{ var }}, na.rm = TRUE),
      median = median({{ var }}, na.rm = TRUE),
      min = min({{ var }}, na.rm = TRUE),
      max = max({{ var }}, na.rm = TRUE),
      q25 = quantile({{ var }}, 0.25, na.rm = TRUE), # first quartile
      q75 = quantile({{ var }}, 0.75, na.rm = TRUE), # third quartile
      sd = sd({{ var }}, na.rm = TRUE),
      n = length(na.omit({{ var }}))
    ) %>%
    mutate(IQR = q75 - q25)
}

#a simple qq test function. It probably doesn't actually save me any time, honestly.
#usage: qqTest(data, sample, "title")
qqTest <- function(data, sample, title = "title") {

  require(tidyverse)

  data %>%
  ggplot(aes(sample = {{ sample }})) +
  stat_qq() +
  stat_qq_line(
    col = "#ecd467",
    size = 0.8) +
    labs(
      title = title,
      x = "Theoretical Quantiles",
      y = "Sample Quantiles"
      )
}

# Function to summarize variables for multiple groups
# Usage: summarize_group_var(dataset, group_var, var)
summarize_group_var <- function(dataset, group_var, var) {
  require(tidyverse)

  dataset %>%
    group_by({{ group_var }}) %>%
    summarise(
      mean = mean({{ var }}, na.rm = TRUE),
      median = median({{ var }}, na.rm = TRUE),
      min = min({{ var }}, na.rm = TRUE),
      max = max({{ var }}, na.rm = TRUE),
      q25 = quantile({{ var }}, 0.25, na.rm = TRUE), # first quartile
      q75 = quantile({{ var }}, 0.75, na.rm = TRUE), # third quartile
      sd = sd({{ var }}, na.rm = TRUE),
      n = length(na.omit({{ var }}))
    ) %>%
    mutate(IQR = q75 - q25)
}

#faceted boxplot function with one box per group
#usage: facBox1(data, x, y, fill, facet, title)
facBox1 <- function(data, x, y, fill, title) {

  require(tidyverse)

  ggplot(data, aes(x = {{ x }}, y = {{ y }}, fill = {{ fill }})) +
    geom_boxplot() +
    labs(title = title)
}

```

Plotly Functions

```{r}

# Test histogram function using Plotly
# Example usage:
# plot <- histogramCurvePlotly(data, data$x, "Title", "X-axis Label")
# plot
library(plotly)

histogramCurvePlotly <- function(data, x, title, x_title) {
    ggthemr("flat dark")

  p <- plot_ly(data, x = ~x, type = "histogram", histnorm = "probability density", showlegend = FALSE, nbinsx = 40) %>%
    layout(
        title = title,
        xaxis = list(title = x_title),
        yaxis = list(title = "Density"),
        font = list(color = "#ffffff"), # Set text color to white
        paper_bgcolor = "#1e1e1e",      # Set background color
        plot_bgcolor = "#1e1e1e"        # Set plot area color
    )

  mean_val <- mean(x)
  sd_val <- sd(x)
  
  x_range <- seq(min(x), max(x), length.out = 100)
  curve_vals <- dnorm(x_range, mean = mean_val, sd = sd_val)
  
  p <- add_lines(p, x = x_range, y = curve_vals, type = "scatter", mode = "lines", line = list(color = "#ecd467"))
  
  return(p)
}


#faceted boxplot written in plotly
facBoxPlotly <- function(data, x, y, fill, title) {

  require(ggthemr)
  require(plotly)

  ggthemr("flat dark")

  p <- plot_ly(data, x = x, y = y, color = fill, type = "box") %>%
    layout(
        title = title,
        font = list(color = "#ffffff"), # Set text color to white
        paper_bgcolor = "#1e1e1e",      # Set background color
        plot_bgcolor = "#1e1e1e"        # Set plot area color
    )

  return(p)
}


hist.dodge.plotly <- function(data, x, group, title = "Insert Title") {
  require(ggplot2)
  require(ggh4x)

  gg <- ggplot(data, aes(x = x, fill = group)) +
    geom_histogram(aes(y = after_stat(density)),
                   bins = 45,
                   alpha = 0.8,
                   color = "#353535") +
    stat_theodensity(col = "#ecd467",
                     size = 0.9,
                     linetype = "dashed") +
    labs(title = title, x = "Group", y = "Frequency") +
    facet_wrap(group) +
    theme(
      plot.background = element_rect(fill = "#1e1e1e"),
      panel.background = element_rect(fill = "#1e1e1e"),
      axis.text = element_text(color = "#ffffff"),
      axis.title = element_text(color = "#ffffff"),
      strip.text = element_text(color = "#ffffff"),
      strip.background = element_rect(fill = "#1e1e1e")
    )

  ggplotly(gg, tooltip = "text") %>%
    layout(
      font = list(color = "#ffffff"),  # Set text color to white
      paper_bgcolor = "#1e1e1e",      # Set background color
      plot_bgcolor = "#1e1e1e"        # Set plot area background color
    )
}


#regular boxplot written in plotly
PlotlyBox <- function(data, y, title) {

  require(ggthemr)
  require(plotly)

  ggthemr("flat dark")

  p <- plot_ly(data, y = y, type = "box") %>%
    layout(
        title = title,
        font = list(color = "#ffffff"), # Set text color to white
        paper_bgcolor = "#1e1e1e",      # Set background color
        plot_bgcolor = "#1e1e1e"        # Set plot area color
    )

  return(p)
}

```

## Load and Cleaup Data

```{r}

library(knitr)
library(dplyr)

load("acupuncture_data_reduced.RData")
acupuncture <- data #creating table with more descriptive name while leaving original untouched

#creating variable for change in headache severity between baseline and 3-month follow-up (pk2 - pk1)
#changing certain columns to factors
acupuncture <- acupuncture %>% #making data set with more descriptive label
    dplyr::mutate(Severity_Change = (pk2 - pk1)) %>%
    dplyr::mutate(group_f = factor(group, levels = c(0, 1), labels = c("control", "treatment"))) %>%
    dplyr::mutate(across(c(id, sex, migraine, acupuncturist, practice_id, group), as.factor))

knitr::kable(head(acupuncture), caption = "Table 1. Select Observations From the Dataset")

```

# Part 1: One-Sample

## Descriptives

### Summary Statistics

```{r}

#creating specific data frames for change in severity
changeSummary <- summarize.var(acupuncture, Severity_Change)
kable(changeSummary, caption = "Table 2. Summary Statistics of Severity Change (n excludes NAs)")
sevChange <- acupuncture %>%
  select(pk1, pk2, Severity_Change, group_f) %>%
  mutate(n_pk1 = c(length(na.omit(pk1)))) %>%
  mutate(n_pk2 = c(length(na.omit(pk2))))
sevChangeNA <- sevChange %>% na.omit #creating frame without NAs

```

## Plots

```{r, dpi = 150, warning = FALSE}

hist.summarycurve(sevChangeNA, "Severity_Change", "Fig 1. Histogram of Change in Headache Severity", "Change (combined groups)")

qqTest(sevChangeNA, Severity_Change, "Fig 2. QQ-plot of Average Medicare Payments")

#sevChangeNA %>% 
#  ggplot(aes(y = Severity_Change)) +
#  geom_boxplot() +
#  labs(
#    title = "Fig 2. Boxplot of Severity Change",
#    y = "Change"
#  )

```

The distribution appears normally distributed with a **mean of -4.7, median of -4,**
**standard deviation of 11.7, and IQR of 10.4.** The data are slightly skewed with a **skewness of -0.17.**

## Paired t test

```{r}

ttest1 <- t.test(acupuncture$pk2, acupuncture$pk1, paired = TRUE, conf.level = 0.99) %>%
  broom::glance()
kable(ttest1, caption = "Table 3. Two-Sided Paired T-Test of Change in Headache Severity. NAs excluded. N pk1 = 401. N pk2 = 326")

```

A paired test was used as we are
comparing headache severity in the same group of people before and after treatment/control.

The null hypothesis was that the change in headache severity would be 0, with the alternative
being that the change would not be 0.

The test resulted in a **T = -7.2599** and **p = 2.877e-12.** 
The two-sided confidence interval is **(-6.3974, -3.0321).**

With the low p-value and lack of 0 appearing in the 99% confidence interval, we reject the null
hypothesis null hypothesis and conclude that there is a significant difference in the severity
of headaches over the three months to follow-up in the combined control and treatment groups.

## Signed rank test

```{r}

wilcox1 <- wilcox.test(acupuncture$pk2, acupuncture$pk1, conf.level = 0.99, conf.int = TRUE) %>%
  broom::glance()
kable(wilcox1, caption = "Table 4. Wilcoxon Signed-Rank Test of Change in Headache Severity")

```

A Wilcoxon Signed-Rank Test comparing the medians of headache severity from baseline to
the three month followup yielded **p = 1e-07.** 
We meet the conditions for the test as the this is a random sample and the distributions are **symmetrical.**

The null hypothesis is that the medians are the same; the alternative is that the medians
are not the same.

With the calculated p-value, we **reject the null hypothesis** that the medians are the same. There is sufficient
evidence that there is a significant difference in the severity of headaches between the first measurement at at the 
three month followup.

## Summary of part 1 

Based on the findings, we are able to conclude that there was a statistically significant chage in the severity of headaches between initial
report and followup, with a mean change of -4.7. While the p value changed depending on the test, both parametric and nonparametric tests
yielded a p value below our alpha of 0.01.

# Part 2: Two Sample

## Descriptives

### Summary Statistics

```{r}

changeSummaryGroup <- summarize_group_var(acupuncture, group_f, Severity_Change)
kable(changeSummaryGroup, caption = "Table 5. Summary Statistics of Headache Severity Change by Group")

```

The data for both groups appear symetric. The distribution of the treatment group contains a higher maximum and lower minimum.

## Plots

```{r, dpi = 150}

facBox1(sevChangeNA, Severity_Change, group_f, group_f, "Fig 3. Boxplots of Change in Severity by Group")

hist.dodge(sevChangeNA, sevChangeNA$Severity_Change, sevChangeNA$group_f, "Fix 4. Histograms of Change in Severity by Group")

```

## Two-sample t test

```{r}

ttest2 <- t.test(Severity_Change ~ group_f, data = sevChangeNA, var.equal = FALSE) %>%
  broom::glance()
kable(ttest2, caption = "Table 6. Two-Sided Two-Sample T-Test Comparing Change in Headache Severity by Treatment Group")

```

A two-sample t test was performed without assuming equal variance. The null hypothesis was that there would not be
a difference in mean headache severity change between treatment and control groups, and the alternative was that there
would be a difference. We meet the conditions for this test as we have random samples, two independent groups, and a sufficiently large n.
This yielded **T = 2.9006 and p = 0.004.** The 95% confidence interval is **(1.1923, 6.2185).**

Based on the test, we **reject the null hypothesis** that the difference in mean severity change = 0 and **fail to reject the alternative**
that they are not equal. We conclude that there is evidence that the treatment resulted in a change different from the placebo.

## Wilcoxon rank sum test


```{r}

wilcox2 <- wilcox.test(Severity_Change ~ group_f, data = sevChangeNA) %>%
  broom::glance()
kable(wilcox2, caption = "Table 7. Wilcoxon Rank Sum Test of Change in Headache Severity by Treatment Group")

```

A Wilcoxon rank sum test comparing the median change in headache severity between the two groups yielded **p = 0.0012.**
The conditions for the test were met as we used random samples and two independent groups with similar distributions.
From the results of the test, we **reject the null hypothesis** that the medians are equal and **fail to reject the alternative**
that they are not equal. We conclude that the treatment resulted in a difference in change in headache severity compared to the placebo.

## Summary of part 2

Based on our findings, we can conclude that there is likely a difference in change in headache severity by the followup date between the treatment group
and the control group. While the nonparametric test yielded a higher p value than the parametric test, both tests were within alpha = 0.05.

# Session Information

```{r}
sessionInfo()
```

# Plotly Graphs that I was having "fun" making.

```{r}

HistChangeCombined.Plotly <- histogramCurvePlotly(sevChangeNA, sevChangeNA$Severity_Change, "Fig 1. Histogram of Change in Severity (combined groups)", "X-axis Label")
HistChangeCombined.Plotly

qqCombined.Plotly <- qqTest(sevChangeNA, Severity_Change, "Fig 2. QQ-plot of Average Medicare Payments")
ggplotly(qqCombined.Plotly)

BoxCombined.Plotly <- PlotlyBox(sevChangeNA, sevChangeNA$Severity_Change, "Fig 3. Boxplot of Severity Change")
BoxCombined.Plotly

######

BoxGroups.Plotly <- facBoxPlotly(sevChangeNA, sevChangeNA$Severity_Change, sevChangeNA$group_f, sevChangeNA$group_f, "Fig 4. Boxplots of Change in Severity by Group")
BoxGroups.Plotly

HistGroups.Plotly <- hist.dodge.plotly(sevChangeNA, sevChangeNA$Severity_Change, sevChangeNA$group_f, "Fig 5. Histograms of Change in Headache Severity by Group")
ggplotly(HistGroups.Plotly)



```