---
title: "M4 Lab"
author: "Matthew Præstgaard"
date: "due 11/12/2023"
output:
  rmdformats::downcute:
    self_contained: true
    default_style: "dark"
    downcute_theme: "default"
    code_folding: "show"
  svglite:
    fig.retina: 2
---

```{r,echo = FALSE, message = FALSE, warning = FALSE}
require(knitr)
knitr::opts_chunk$set(error = TRUE)
```

```{r, message = FALSE}

library(tidyverse)
library(car)
library(emmeans)
library(ggthemr)
    ggthemr("flat dark")
library(markdown)
library(rmdformats)
library(plotly)
library(moments)
library(gridExtra)
library(ggh4x)
library(jtools)
#library(kableExtra)

```

Functions

```{r}

# Function to summarize variables with optional grouping
# Usage: summarize.var(dataset, var, group_var (optional))
summarize.var <- function(dataset, var, group_var = NULL) {
  require(tidyverse)

  result <- dataset %>%
    {if (!is.null(group_var)) group_by({{ group_var }}) else .} %>%
    summarise(
      mean = mean({{ var }}, na.rm = TRUE),
      median = median({{ var }}, na.rm = TRUE),
      min = min({{ var }}, na.rm = TRUE),
      max = max({{ var }}, na.rm = TRUE),
      q25 = quantile({{ var }}, 0.25, na.rm = TRUE), # first quartile
      q75 = quantile({{ var }}, 0.75, na.rm = TRUE), # third quartile
      sd = sd({{ var }}, na.rm = TRUE),
      n = length(na.omit({{ var }}))
    ) %>%
    mutate(IQR = q75 - q25)

  return(result)
}


# Faceted boxplot function with one box per group
# Usage: ggBoxplots(data, x, fill, title)
# Example usage:
# ggBoxplots(your_data_frame, x_column_name, fill_column_name, "Your Title")
ggBoxplots <- function(data, x, fill, title = "title") {
  require(tidyverse)

  ggplot(data, aes(x = {{ x }}, y = "", fill = {{ fill }})) +
    geom_boxplot() +
    labs(title = title) +
    facet_wrap(as.formula(paste0("~", as.name(deparse(substitute(fill))))), scales = "free")
}


# ggplot histogram function with optional faceting variable
# Usage: gg.hist(data, data$x, bins (optional), title (optional), group (optional))
gg.hist <- function(data, x, bins = 20, title = "Insert Title", x_lab = "x", group = NULL) {
  require(ggplot2)
  require(ggh4x)

  gg <- ggplot(data, aes(x = {{ x }})) +
    geom_histogram(aes(
        y = after_stat(density)),
      bins = bins,
      alpha = 0.8,
      color = "#353535"
    ) +
    stat_theodensity(
      col = "#ecd467",
      size = 0.9,
      linetype = "dashed"
    ) +
    labs(
      title = title,
      x = x_lab,
      y = "Frequency"
    )

  if (!is.null(group)) {
    gg <- gg + aes(fill = {{ group }}) +
      facet_wrap({{ group }})
  }

  return(gg)
}



qqTest <- function(data, sample, title = "title", facet_var = NULL) {
    require(tidyverse)

    gg <- ggplot(data,
        aes(sample = {{ sample }})) +
        stat_qq() +
        stat_qq_line(
            size = 0.8,
        ) +
        labs(
            title = title,
            x = "Theoretical Quantiles",
            y = "Sample Quantiles"
        ) +
        guides(color = "none")

    if (!is.null(facet_var)) {

        # Extract colors from facet_var variable
        colors <- pull(data, {{ facet_var }})

        gg <- gg + aes(color = factor(colors)) + facet_wrap(as.formula(paste("~", facet_var)))
    }

    return(gg)
}



#density plots with optional grouping variable
#can be faceted or together by specifying either group or color
#usage: gg.density(data, sample, group = "group" (optional), color = "NULL" (optional), title = "title" (optional) )
gg.density <- function(data, sample, group = NULL, color = NULL, title = "Insert Title") {

  require(ggplot2)

  plot <- ggplot(data, aes(x = {{ sample }})) +
    geom_density() +
    labs(
      title = title,
      y = "Frequency"
    )

  if (!is.null(group)) {

    colors <- pull(data, {{ group }})

    plot <- plot + aes(color = factor(colors)) +
      facet_wrap(as.formula(paste("~", group))) +
      labs(color = "Group")
  }

  if (!is.null(color)) {

    colors <- pull(data, {{ color }})

    plot <- plot + aes(fill = factor(colors), color = factor(colors), alpha = 0.02) +
      labs(fill = "Group") +
      geom_density(size = 1) +
      guides(color = "none", alpha = "none")
  }

  return(plot)
}

#ggplotRegression
#use to plot linear model with summary statistics and error
#Usage: ggplotRegression(data)
ggplotRegression <- function(fit, title = "title") {

require(ggplot2)

ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
  geom_point() +
  stat_smooth(
    method = "lm",
    geom = "smooth") +
  labs(
    title = title,
    subtitle = paste("Adj R2 = ", signif(summary(fit)$adj.r.squared, 5), 
                    "Intercept =", signif(fit$coef[[1]], 5 ), 
                    " Slope =", signif(fit$coef[[2]], 5), 
                    " P =", signif(summary(fit)$coef[2,4], 5)))
}

```

# Part 1: Tumor Growth

## Data Management

### Load Data

```{r}

# load workspace depending on which OS being used
sysname <- Sys.info()["sysname"]

if(sysname == "Linux") {
  load("/home/matthew/rclone-gdrive/Grad School/Fall23/BIOST_2241_Intro/M4/Lab4/tumor_growth.RData")
} else if(sysname == "Windows") {
  load("X:/My Drive/Grad School/Fall23/BIOST_2241_Intro/M4/Lab4/tumor_growth.RData")
}

```

```{r}

# add factor variable for treatment labels
tumordat <- tumor %>%
    dplyr::mutate(
        Group_f = factor(Group, levels = c(1, 2, 3, 4),
        labels = c("control", "drug_only", "radiation_only", "drug_radiation")))
kable(head(tumordat), caption = "Table 1. Select Data")

```

## Descriptive Statistics

```{r}

# numerical summary by group
sum1 <- summarize.var(tumordat, Group_f, Size)
kable(head(sum1), caption = "Table 2. Summary Statistics")

#making table with just group and size variables
group_size <- tumordat %>%
    dplyr::select(Size, Group_f) %>%
    data.frame()

```

## Plots

```{r, warning = FALSE}
# plots
gg.hist(group_size, Size, title = "Fig 1. Histograms of Size by Group", group = group_size$Group_f) %>%
  ggplotly()

gg.density(group_size, Size, color = "Group_f", title = "Fig 2. Density Plots of Size by Group") %>%
  ggplotly()

#hist1 + labs(subtitle = "means (clockwise from top left) = 2067.29, 1787.75, 1890.32, 1130.64")


```

The data do not appear to deviate too far from a normal distribution, although the sample size per treatment group is very small.

## ANOVA

As the sample sizes are quite small, qq-plots were created to assess normality. The data appear to
follow the theoretical quantiles somewhat, but the test still should not be taken as
conclusive as it would be if N were larger.

```{r}

# Check conditions

# sample sizes table
sum1 %>% dplyr::select(Group_f, n) %>% kable(Caption = "Table 3. Sample Sizes of Groups")

```


```{r, include = FALSE}

# qqplots for each group if sample sizes are small
qq1 <- qqTest(group_size, Size, "Fig 3. QQ-plots of Size by Group", "Group_f")

```

```{r}

ggplotly(qq1)

```


```{r}
# run anova model

# estimate group means with lm formula
model <- lm(Size ~ Group_f, data = tumordat)

# display ANOVA output
Anova(model)

# multiple-testing correction
bonf.model <- emmeans(model, pairwise ~ "Group_f", adjust = "bonferroni")

# show each pairwise comparison
knitr::kable(bonf.model$contrasts)

```

For the ANOVA test, the **null hypothesis** was that none of means differed. The **alternative hypothesis** was that there was
at least one mean that differed from the rest of the groups.
The p-value from the initial ANOVA test was **0.003.** Pairwise comparison tests showed the largest difference in to be 
between the control and combined radiation-drug therapy groups (p = 0.0038).
We can conclude from this that the most effective treatment is likely a combination drug and radiation therapy.



## Kruskal-Wallis


```{r}
# check conditions

# run test
kruskalModel <- kruskal.test(Size ~ Group_f, data = tumordat)
kruskalModel

```

The hypothesis for this test were that none of the group medians would differ **(null)** and that at least one of the medians would differ **(alternative).**
We assume that we have random samples in this and that they are independent. Looking at the qq-tests shows them to be similar in distribution.
With **p = 0.023** we can conclude from the test that at least one of the group medians is different from the rest. This is in line with the results from the ANOVA test.



# Part 2: Intubation

## Data Management 


```{r}
# load in data and filter
sysname <- Sys.info()["sysname"]

if(sysname == "Linux") {
  load("/home/matthew/rclone-gdrive/Grad School/Fall23/BIOST_2241_Intro/M4/Lab4/Laryngoscope.Rdata")
} else if(sysname == "Windows") {
  load("X:/My Drive/Grad School/Fall23/BIOST_2241_Intro/M4/Lab4/Laryngoscope.Rdata")
}
# keep only successful first intubation attempts, BMI, and Ease
laryClean <- laryngoscope %>%
  select(attempt1_S_F, BMI, ease) %>%
  filter(attempt1_S_F == 1)

```

## Descriptive Statistics

```{r}

# numerical summary
larySum <- summarize.var(laryClean, BMI)
knitr::kable(larySum)

# plots
gg.hist(laryClean, BMI, title = "Fig 4. Histogram of BMI") %>%
  ggplotly()

gg.hist(laryClean, ease, title = "Fig 5. Histogram of Ease of Intubation") %>%
  ggplotly()

qqBMI <- qqTest(laryClean, BMI, title = "Fig 6a. QQ-plot of BMI")
qqEase <- qqTest(laryClean, ease, title = "Fig 6b. QQ-plot of Ease of Intubation")
qqlary <- grid.arrange(qqBMI, qqEase)

```

The distribution of ease of intubation seems vaguely bimodal, while the distribution of BMI appears normal with a slight right skew.
The qq-tests for each suggest that they are fairly normally distributed.

## Linear Regression  

```{r}
# fit the model
fit <- lm(ease ~ BMI, data = laryClean)

ggplotRegression(fit, title = "Fig 6. Regression Plot of Ease of Intubation vs BMI") %>%
  ggplotly()

# estimated coefficients, t-tests, and global F-test
#summary(fit)
summ(fit) #this may not be visible in dark mode

#confidence interval
confint(fit)

# predict the value for BMI = 35

```


The equation for the linear model of ease of intubation vs BMI is **y = -0.53047x + 62.781.**
This means that for every 1 point increase in BMI, the ease of intubation is expected to decrease by aprox. 0.53.
With an adjusted R2 of ~0, it can be said that none of the variance in ease of intubation is attributable to BMI.
The p-value for the slope is **0.37,** indicating that we have no evidence that BMI has a linear relationship with ease of intubation.



# Session Information

```{r}
sessionInfo()
```